Global code motion of congruent instructions

### 2017-04-17 paper.tex
conclusion

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on overall compile time on \SPEC{}. GCM enables more inlining and
exposes more loop invariant code motion opportunities in majority of the
benchmarks. We have also seen execution time improvements in a few of \SPEC{}
benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
\SPEC{} bencmarks suite when compiled for X86\_64. This is an optimistic
algorithm in the sense that it considers all identical computations in a
function as potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of code-motion, and how
code-motion increases instruction level parallelism thereby enabling more out of
order execution on modern architectures.
\end{abstract}

@@

\begin{verbatim}
void doCodeSinking()
  For each value number VN with 2 or more instructions:
    if there are two instructions with same immediate successor as post-dominator
      if both are only used in the same PHI-node of the successor
        if both have dependencies that are available or can be made available at the sink point
           mark VN as sinkable

  For each pair I1, I2 of sinkable instructions:
    Move the first one I1 to the sink point which is just after all the PHI-nodes
    in the post-dominator
    update the use of PHI with I1
    Remove I2 and PHI
    if I1 is a memory reference:
      update MemorySSA of I1 and others to point to the MemorySSA reference of I1
      remove MemorySSA reference of all others which were deleted
    update statistics

@@

While benchmarking \SPEC{} we see both increase as well as decrease in
the codesizes of the final binaries. Since the pass runs early, it affects many
optimizations which rely on the number of instructions, lenth of the use-def
chain, and other metrics. For instance, the inliner is impacted by a decrease in
the number of instructions in the caller and callee, as its heuristics estimate
the size of functions to be inlined. Various code-size metrics are shown in
Table~\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
by 35.38\%. In this benchmark 3 more functions got inlined (15 as compared to
12) and because of that 10 more vector instructions got generated (81 vs. 71), 3
calls got hoisted/sunk as (compared to 0), one loop got unswitched (compared to
0), 6 high latency machine instructions got hoisted out of loop, 59 (compared to
30) machine instructions got hoisted out of loop, 70 (compared to 39) machine
instructions were sunk.

@@

\section{Conclusion and Future Work}
\label{sec:future-work}
We have presented the \GVN{} based code hoisting algorithm. The primary goal is to
reduce the code size but it benefits performance in some cases as well. To
preserve performance and not hoist too much we have implemented several cost
models described in Section~\ref{sec:cost-models}. Since those cost models
depend on a set of thresholds, it requires tuning, as such, we used
representative benchmarks to tune them.

In general it is a good idea to start with lower ranked expression first such
that maximum hoisting can happen in one iteration, however, current
implementation does not rank the expressions and iteratively finds a fixed point
when no more candidates are available. Even this implementation converges
quickly and no significant compile time regression have been observed because of
code hoisting pass. This is not the most optimal approach and results in
multiple data structures to be recomputed. This can be improved by ranking the
computations \cite{rosen1988global}. Also \GVN{}-hoist runs very early in the
pass pipeline, it will be good to evaluate the codesize/performance impact when
it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.

With the implementation of code-hoisting in \LLVM{}, the passes which rely on the
code-size/instruction-count to make optimization decisions needs to be
revisited. The first candidate would be the inliner. We have seen different
inlining decisions in Table~\ref{tab:code-size}, before and after
code-hoisting was enabled.  Since inliner has several magic numbers tuned for
the previous pass layout, it would need some improvement.

>>>>

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on overall compile time on \SPEC{}. GCM enables more inlining and
exposes more loop invariant code motion opportunities in majority of the
benchmarks. We have also seen execution time improvements in a few of \SPEC{}
benchmarks viz. mcf and sjeng, moreover, register spills reduced by 2\% on the
\SPEC{} bencmarks suite when compiled for X86\_64. This is an optimistic
algorithm in the sense that it considers all congruent instructions in a
function as potential candidates. We make an extra effort to hoist candidates by
partitioning the potential candidates in a way to enable partial hoisting in
case common hoisting points for all the candidates cannot be found. We also
formalize how register pressure will reduce as a result of code-motion, and how
code-motion increases instruction level parallelism thereby enabling more out of
order execution on modern architectures.
\end{abstract}

@@

\begin{verbatim}
void doCodeSinking()
  For each value number VN with 2 or more instructions:
    if there are two instructions with same immediate successor as post-dominator
      if both are only used in the same PHI-node (PN) of the successor
        if both have dependencies that are available or can be made available at the sink point
           mark VN as sinkable

  For each pair I1, I2 of sinkable instructions:
    Move the first one I1 to the sink point which is just after all the PHI-nodes
    in the post-dominator
    update all the uses of PHI (PN) with I1
    Remove I2 and PN
    if I1 is a memory reference:
      update MemorySSA of I1 and others to point to the MemorySSA reference of I1
      remove MemorySSA reference of all others which were deleted
    update statistics

@@

While benchmarking \SPEC{} we see both increase as well as decrease in
the codesizes of the final binaries. Since the pass runs early, it affects many
optimizations which rely on the number of instructions, lenth of the use-def
chain, and other metrics. For instance, the inliner is impacted by a decrease in
the number of instructions in the caller and callee, as its heuristics estimate
the size of functions to be inlined.

@@

\section{Conclusion and Future Work}
\label{sec:future-work}
We have presented the global code motion of congruent computations in SSA
form. We saw that it improves inlining and LICM opportunities, reduces register
spills, it also improves performance in some cases. To preserve performance and
not hoist too much we have register pressure aware cost models described in
Section~\ref{sec:cost-models}. A part of this implementation is already merged
in LLVM trunk as GVNHoist.cpp. We will post the GCM implementation for review.

There are a few improvements that we would like to make. First would be to
integrate this with \LLVM{'}s new GVN-PRE implementation such that more
congruent instructions can be scheduled. This would also expose more
redundancies to the GVN-PRE.  In general it is a good idea to start with lower
ranked expression first such that maximum hoisting can happen in one iteration,
however, current implementation does not rank the expressions and iteratively
finds a fixed point when no more candidates are available. Even this
implementation converges quickly and no compile time regression have been
observed because of code hoisting pass. This is not the most optimal approach
and results in multiple data structures to be recomputed. This can be improved
by ranking the computations \cite{rosen1988global}. Also, since GCM runs very
early in the pass pipeline, it will be good to evaluate the codesize/performance
impact when it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.

With the implementation of GCM in \LLVM{}, the passes which rely on the
code-size/instruction-count to make optimization decisions needs to be
revisited. The first candidate would be the inliner. We have seen different
inlining decisions in Table~\ref{tab:code-size}, before and after code-hoisting
was enabled.  Since inliner has several magic numbers tuned for the previous
pass layout, it would need some improvement.
